A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2023-12-27 paper/paper.tex
Finish typesetting the proof of determinism

\begin{figure}[h]
\begin{equation*}
  \begin{split}
    [[InterpR i A S]] &:= [[ Interp F i A S  ]] \\
    [[F i]] &:= [[exists S , InterpR i A S]]
  \end{split}
\end{equation*}
\caption{Logical relation for all universe levels}
\label{fig:logrelrec}
\end{figure}
To tie the knot and obtain an interpretation of all universe levels,
we define in Figure~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
using the well-foundedness of the strict less than relation on natural
numbers.
The judgment $[[InterpR i A S]]$ now reads that the type $[[A]]$ is a
level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
$[[S]]$. The way it's constructed enables us to reason about its
property by nested induction over the less than relation and the
structure of the relation $[[Interp I i A S]]$, though for the
majority of the cases we do not need any information about the
parameterized function $[[I]]$, where the property about $[[InterpR i
A S]]$ follows as a corollary of a property about $[[Interp I i A S]]$
with no assumptions imposed on $[[I]]$.

@@

The relation $[[Interp I i A S]]$ satisfies the following inversion
principles.
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
  \item If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
  \item If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
  \item If $[[Interp I i Pi A B S]]$, then there exists $[[S]],[[F]]$ such that:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
  \item If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  We only show the most involved function case; the rest follows a
  similar but simpler pattern. We start by
  inducting over the derivation of $[[Interp I i Pi A B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$, and from the
    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \end{itemize}
    By inverting $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[F]]$ satisfies the desired properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
  \end{description}
\end{proof}

\Rref{I-Red} bakes into the logical relation the backward preservation
property. The following property shows that preservation holds in the
usual forward direction too.
\begin{lemma}[Preservation of the logical relation]
  \label{lemma:interppreservation}
  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
\end{lemma}
\begin{proof}
  We carry out the proof by induction over the derivation of $[[Interp
  I i A S]]$.
  \begin{description}
  \item[\Rref{I-Void}:] There exists some $[[B]]$ such that $[[Void =>
    B]]$. By inverting the derivation of $[[Void => B]]$, $[[B]]$ must
    be $[[Void]]$ and the result trivially follows.
  \item[\Rref{I-Bool, I-Set}:] Similar to the case for \rref{I-Void}.
  \item[\Rref{I-Eq}:] We know that $[[Interp I i a ~ b : A { p | p =>+
      refl , a <=> b  }]]$ and, by inverting the derivation of
    parallel reduction, $[[a => a0]]$, $[[b => b0]]$, $[[A => A0]]$
    for some $[[a0]]$, $[[b0]]$, and $[[A0]]$. Our goal is to show
    that $[[Interp I i a0 ~ b0 : A {p | p =>+ refl, a <=> b}]]$. By
    \rref{I-Eq}, we already know that $[[Interp I i a0 ~ b0 : A {p | p
      =>+ refl, a0 <=> b0}]]$ and therefore it suffices to show that
    the sets $[[{p | p =>+ refl, a <=> b}]]$ and $[[{p | p =>+ refl,
      a0 <=> b0}]]$ are equal. Equivalently, it suffices to show that
    $[[a <=> b]]$ if and only if $[[a0 <=> b0]]$. By definition, from
    $[[a => a0]]$ and $[[b => b0]]$, we derive $[[a <=> a0]]$ and $[[b
    <=> b0]]$. The result then immediately follows from the fact that
    coherence is an equivalence relation
    (Lemma~\ref{lemma:coherenceequiv}).
  \item[\Rref{I-Pi}:]
  \end{description}
\end{proof}

>>>>

\begin{figure}[h]
\begin{equation*}
    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I i]] := [[{A | exists S , InterpR i A S}]]
\end{equation*}
\caption{Logical relation for all universe levels}
\label{fig:logrelrec}
\end{figure}
To tie the knot and obtain an interpretation of all universe levels,
we define in Figure~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
using the well-foundedness of the strict less than relation on natural
numbers (recall that
the parameter $[[I]]$ of $[[Interp I i A S]]$ takes only natural
numbers strictly less than $[[i]]$ as its input).
The judgment $[[InterpR i A S]]$ now reads that the type $[[A]]$ is a
level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
$[[S]]$. The way it's constructed enables us to reason about its
property by nested induction over the less than relation and the
structure of the relation $[[Interp I i A S]]$, though for the
majority of the cases we do not need any information about the
parameterized function $[[I]]$, where the property about $[[InterpR i
A S]]$ follows as a corollary of a property about $[[Interp I i A S]]$
with no assumptions imposed on $[[I]]$.

@@

The relation $[[Interp I i A S]]$ satisfies the following inversion
principles.
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
  \item If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
  \item If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
  \item If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
  \item If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  We only show the most involved function case; the rest follows a
  similar but simpler pattern. We start by
  inducting over the derivation of $[[Interp I i Pi A B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S0]]$.
    We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
    A0 B0 S0]]$. From the
    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
    By inverting $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[F]]$ satisfies the following properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    The properties above are exactly the preconditions needed to apply
    \rref{I-Pi} for $[[Pi A B]]$ to finish off the proof.
  \end{description}
\end{proof}

\Rref{I-Red} bakes into the logical relation the backward preservation
property. The following property shows that preservation holds in the
usual forward direction too.
\begin{lemma}[Preservation of the logical relation]
  \label{lemma:interppreservation}
  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
\end{lemma}
\begin{proof}
  We carry out the proof by induction over the derivation of $[[Interp
  I i A S]]$.
  \begin{description}
  \item[\Rref{I-Void}:] There exists some $[[B]]$ such that $[[Void =>
    B]]$. By inverting the derivation of $[[Void => B]]$, $[[B]]$ must
    be $[[Void]]$ and the result trivially follows.
  \item[\Rref{I-Bool, I-Set}:] Similar to the case for \rref{I-Void}.
  \item[\Rref{I-Eq}:] We know that $[[Interp I i a ~ b : A { p | p =>+
      refl , a <=> b  }]]$ and, by inverting the derivation of
    parallel reduction, $[[a => a0]]$, $[[b => b0]]$, $[[A => A0]]$
    for some $[[a0]]$, $[[b0]]$, and $[[A0]]$. Our goal is to show
    that $[[Interp I i a0 ~ b0 : A {p | p =>+ refl, a <=> b}]]$. By
    \rref{I-Eq}, we already know that $[[Interp I i a0 ~ b0 : A {p | p
      =>+ refl, a0 <=> b0}]]$ and therefore it suffices to show that
    the sets $[[{p | p =>+ refl, a <=> b}]]$ and $[[{p | p =>+ refl,
      a0 <=> b0}]]$ are equal. Equivalently, it suffices to show that
    $[[a <=> b]]$ if and only if $[[a0 <=> b0]]$. By definition, from
    $[[a => a0]]$ and $[[b => b0]]$, we derive $[[a <=> a0]]$ and $[[b
    <=> b0]]$. The result then immediately follows from the fact that
    coherence is an equivalence relation
    (Lemma~\ref{lemma:coherenceequiv}).
  \item[\Rref{I-Pi}:] There exists $[[S]]$ and $[[F]]$
    such that:
    \begin{itemize}
    \item $[[Interp I i Pi A B { b | forall a, (# a in S implies b a in F
        a #) }]]$
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    There exists some $[[A0]]$ and $[[B0]]$ such that $[[A => A0]]$ and
    $[[B => B0]]$. Our goal is to show that $[[Interp I i Pi A0 B0 { b | forall a, (# a in S implies b a in F
      a #) }]]$. By \rref{I-Pi}, it suffices to show:
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \end{itemize}
    Since $[[A => A0]]$ and $[[B {a} => B0 {a}]]$ forall $[[a]]$, the
    latter of which follows from Lemma~\ref{lemma:parsubst}, the above
    conditions follow immediately from the induction hypothesis.
  \item[\Rref{I-Red}:] There exists $[[B0]]$ and $[[S]]$ such that $[[A => B0]]$ and
    $[[Interp I i B0 S]]$. Given an arbitrary $[[B]]$ such that $[[A
    => B]]$, our goal is to show that $[[Interp I i B S]]$. By the diamond
    property of parallel
    reduction (Lemma~\ref{lemma:pardiamond}), there exists some term
    $[[C]]$ such that $[[B0 => C]]$ and $[[B => C]]$. By the induction
    hypothesis, we derive $[[Interp I i C S]]$ from $[[Interp I i
    B0 S]]$. By \rref{I-Red} and
    $[[B => C]]$, we conclude that $[[Interp I i B S]]$.
  \end{description}
\end{proof}
From Lemma~\ref{lemma:interppreservation} and \rref{I-Red}, we can easily
derive the following corollary that two coherent types have the same
interpretation.
\begin{corollary}[Logical Relation for Coherent Types]
  \label{lemma:logrelcoherence}
  If $[[Interp I i A S]]$ and $[[A <=> B]]$, then $[[Interp I i B S]]$.
\end{corollary}
Next, we show that $[[Interp I i A S]]$ is in fact a partial function.
\begin{lemma}[Logical relation is functional]
  \label{lemma:logreldeter}
  If $[[Interp I i A S0]]$ and $[[Interp I i A S1]]$, then $[[S0 = S1]]$.
\end{lemma}
\begin{proof}
  We start by inducting over the derivation of the first premise $[[Interp I i A
  S0]]$.
  \begin{description}
  \item[\Rref{I-Void}:] We know that $[[Interp I i Void
    emptyset]]$. Given $[[Interp I i Void S1]]$, our goal is to show that
    $[[emptyset = S1]]$. This is immediate by applying the $[[Void]]$ case of
    Lemma~\ref{lemma:interpinv} to $[[Interp I i Void S1]]$.
  \item[\Rref{I-Bool, I-Eq, I-Set}:] Similar to the \rref{I-Void} case
    by applying the matching case of Lemma~\ref{lemma:interpinv}
    to $[[Interp I i A S1]]$.
  \item[\Rref{I-Pi}:] There exists $[[S]]$ and $[[F]]$
    such that:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    Our goal is to show that given $[[Interp I i Pi A B S1]]$, we have
    $[[S1 = { b | forall a, (# a in S implies b a in F
      a #) }]]$.
    By the function case of Lemma~\ref{lemma:interpinv}, there exists
    some $[[S0]]$ and $[[F0]]$ such that:
    \begin{itemize}
    \item $[[Interp I i A S0 ]]$
    \item $[[F0 in S0 -> PowerSet STm]]$
    \item
      $[[forall a, (# a in S0 implies Interp I i B { a } F0 a #)]]$
    \item
      $[[S1 = { b | forall a, (# a in S0 implies b a in F0 a #) }]]$
    \end{itemize}
    It suffices to show that $[[S = S0]]$ and $[[F = F0]]$. The
    equality $[[S = S0]]$ immediately follows from the induction
    hypothesis since $[[Interp I i A S]]$ and $[[Interp I i A
    S0]]$. Therefore, functions $[[F]]$ and $[[F0]]$ have the same
    domain and codomain and thus it suffices to show that forall $[[a in S]]$,
    $[[F a = F0 a]]$. Suppose $[[a in S]]$, we must have $[[Interp I
    i B {a} F a]]$ and $[[Interp I i B {a} F0 a]]$ from the two
    $\forall$-quantified statements above. The equality $[[F a = F0 a]]$ then
    immediately follows from the induction hypothesis.
  \item[\Rref{I-Red}:] There exists some $[[B]]$ such that $[[A =>
    B]]$ and $[[Interp I i B S0]]$. Our goal is to show that given
    $[[Interp I i A S1]]$, we have $[[S0 = S1]]$. By
    Lemma~\ref{lemma:interppreservation}, from $[[Interp I i A S1]]$
    and $[[A => B]]$, we have $[[Interp I i B S1]]$. From the
    induction hypothesis, we can conclude that $[[S0 = S1]]$ since
    $[[Interp I i B S0]]$ and $[[Interp I i B S1]]$.
  \end{description}
\end{proof}
